---
title:  "Peer Assessment  Assignment 1 - Analyzing “quantified self” Movement Data"
---

```{r, warnoff, echo=FALSE}
        options(warn=-1)
```


### **Question 1 - What is mean total number of steps taken per day?**

#### *Read CSV File and Process Data*
```{r, processdata, echo=TRUE}
        alldata <- read.csv("activity.csv", stringsAsFactors=FALSE, na.strings=c("NA", "NULL"))
        alldata <- data.frame(alldata)
        totsteps<- aggregate(steps ~ date, alldata, sum, na.action=na.omit)
        colnames(totsteps) <- c("day", "steps")
```

#### *Table of Total Steps per Day*

```{r, totstepstable, echo=TRUE}
        library(knitr)
        kable(totsteps, col.names=c("date", "total steps"))
```

#### *Histogram of Total Steps per Day*

```{r, totstepshistogram, echo=TRUE}

        library(ggplot2)

        ptitle <- paste(totsteps$day[1], " to ", totsteps$day[length(totsteps$day)])
        ggplot(totsteps,aes(x=steps)) +
                geom_histogram(binwidth=500, fill="lightgreen") +
                ggtitle(toString(ptitle)) +
                xlab("Steps per Day (binwidth = 500)") +
                ylab("Number of People")
                        
```


#### *Tables of Mean and Median of Steps per Day*

```{r, meanmedian, echo=TRUE}
        library(knitr)
        meansteps <- aggregate(steps ~ date, alldata, mean, na.action=na.omit)
        mediansteps <- aggregate(steps ~ date, alldata, median, na.action=na.omit)

        kable(meansteps, digits=1, col.names=c("date", "mean steps"))
        kable(mediansteps, digits=1, col.names=c("date", "median steps"))
```

#### Given the pevalent median value of steps per day being "0" for all days, further 
####    examination indicates users did not begin recording step counts each day
####     until approximately 7am.

# End Question 1

=================================================================

# **Question 2 - What is the average daily activity pattern?**

### *Time Series Plot:  5-minute Intervals (x-axis) versus Average Number of Steps Taken, Averaged Across All Days (y-axis)*

```{r, 5minavesteps, echo=TRUE}
        avesteps <- aggregate(steps ~ interval, alldata, mean, na.action=na.omit)

        plot(avesteps$interval, avesteps$steps, type="l", col="blue", main=toString(ptitle),
             xlab="Daily 5-Minute Intervals", ylab="Average Steps per Interval")
        
        idx <- which(avesteps$steps==max(avesteps$steps))
        points(avesteps$interval[idx], avesteps$steps[idx], pch=19, col="red")

        maxsteps <- round(max(avesteps$steps), digits=0)
        inter <- avesteps$interval[idx]
        legtext <- paste(inter, maxsteps, sep=", ")

        text(avesteps$interval[idx]+5, avesteps$steps[idx], labels=legtext,pos=4)
        legend("topright", c("5-Min Interval, Max Steps"), pch=19,col="red")
```

### *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*
## *interval:  835*



# *Imputing missing values*

```{r,  totalNAvalues, echo=TRUE}
        totalNA <- length(which(is.na(alldata$steps))==TRUE)
```

## *Total NA values:  2304*


### *Removing NA values from "steps"*
#### *Fill NA values with mean of respective interval over all days*

```{r,  fillinNAvalues, echo=TRUE}
        alldata <- read.csv("activity.csv", stringsAsFactors=FALSE)
        alldata <- data.frame(alldata)

        newsteps <- alldata
        newtotsteps<- aggregate(steps ~ interval, newsteps, mean, na.action=na.omit)

        for (i in 1:length(newsteps$steps)){
                idx <- newtotsteps$interval == newsteps$interval[i]
                if(is.na(newsteps$steps[i])){
                        newsteps$steps[i] <- round(newtotsteps$steps[idx])
                }
        }
```


#### *Histogram of Total Steps per Day with NAs Filled*

```{r, newtotstepshistogram, echo=TRUE}

        totsteps<- aggregate(steps ~ date, newsteps, sum)
        colnames(totsteps) <- c("day", "steps")

        ptitle <- paste(totsteps$day[1], " to ", totsteps$day[length(totsteps$day)])
        ggplot(totsteps,aes(x=steps)) +
                geom_histogram(binwidth=500, fill="lightgreen") +
                ggtitle(toString(ptitle)) +
                xlab("Steps per Day (binwidth = 500) with Filled NA") +
                ylab("Number of People") 

```
        

#### *Tables of Mean and Median of Steps per Day with NAs Filled*

```{r, newmeanmedian, echo=TRUE}

        meansteps <- aggregate(steps ~ date, newsteps, mean)
        mediansteps <- aggregate(steps ~ date, newsteps, median)

        kable(meansteps, digits=1, col.names=c("date", "mean steps"))
        kable(mediansteps, digits=1, col.names=c("date", "median steps"))
```

# End Question 2

==================================================

# **Question 3 - Are there differences in activity patterns between weekdays and weekends?**

```{r, weekdayweekend, echo=TRUE}
 
        library(timeDate)

        newsteps$weekday <- isWeekday(newsteps$date, wday=1:5)
        newsteps$weekday <- factor(newsteps$weekday, labels=c("weekend", "weekday"))

        idwkd <- newsteps$weekday == "weekday"
        idwke <- newsteps$weekday == "weekend"

        
        plot(newsteps$interval[idwkd], newsteps$steps[idwkd], xlab="Interval", ylab="Number of Steps", col="blue", type="l", main="Average Steps per 5-Minute Interval")
        lines(newsteps$interval[idwke], newsteps$steps[idwke], col="red")

        legend("topright", c("weekday", "weekend"), lty=c(1,1), lwd=c(2.5,2.5),col=c("blue", "red"), bty="n")
```

#### More steps per 5-minute interval occur in the morning during weekdays, and more steps per 5-minute interval occur in the afternoon on weekends.


# End Question 3